---
title: "World Sustainability Dataset"
author: "RACHEL ATOLAGBE"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggcorrplot")
library(tidyverse)
```



## Importing Libaries

```{r}
# Loading required packages
library(tidyverse)
library(knitr)
library(tibble)
```

## Data Loading

In this section, the dataset is imported into the R environment, to allow for initial exploration and understanding of its fundamental characteristics. 

```{r}
# loading the dataset
WorldSustainability_df <- read_csv("WorldSustainabilityDataset.csv", show_col_types = FALSE)
```

```{r}
# View the structure and types of variables in the dataset
glimpse(WorldSustainability_df)
# to find the minimum and maximum year if the data
min_year <- min(WorldSustainability_df$Year, na.rm = TRUE)
max_year <- max(WorldSustainability_df$Year, na.rm = TRUE)

min_year
max_year
```

```{r}
# identifying numeric variables
numeric_vars <- select_if(WorldSustainability_df, is.numeric)
names(numeric_vars)
cat("Number of numerical variables:", ncol(numeric_vars), "\n")
```

```{r}
# identifying categorical columns
categorical_vars <- select_if(WorldSustainability_df, is.character)
names(categorical_vars)
cat("Number of categorical variables:", ncol(categorical_vars), "\n")
```

## Data Cleaning and Wrangling

```{r}
# Provides a statistical summary of all the variables in the WorldSustainability_df dataset
summary(WorldSustainability_df)

# calculating the total number of missing values in the dataset
sum(is.na(WorldSustainability_df))
```

```{r}
# Cleaning column names by removing indicator codes
colnames(WorldSustainability_df) <- gsub("\\s*-\\s*[A-Z0-9\\.]+\\s*$", "", colnames(WorldSustainability_df))
```

```{r}
# Calculating percentage of missing values for each column
missing_percent <- colSums(is.na(WorldSustainability_df)) / nrow(WorldSustainability_df) * 100

# to view the result as a sorted table (from highest to lowest)
missing_percent_sorted <- sort(missing_percent, decreasing = TRUE)

# to display the result
missing_percent_sorted

```

```{r}
# Selecting the 15 relevant variables for analysis
selected_df <- WorldSustainability_df %>%
  select(
    "Country Name", 
    "Year",
    "Renewable energy consumption (% of total final energy consumption)",
    "Renewable electricity output (% of total electricity output)",
    "Annual production-based emissions of carbon dioxide (CO2), measured in million tonnes",
    "GDP per capita (current US$)",
    "Urban population (% of total population)",
    "Compulsory education, duration (years)",
    "School enrollment, secondary (% gross)",
    "Income Classification (World Bank Definition)",
    "Life expectancy at birth, total (years)",
    "Access to electricity (% of population)",
    "Total natural resources rents (% of GDP)",
    "Population, total",
    "Continent"
  )
```

```{r}
# to view the selected dataset
cat("Number of rows:", nrow(selected_df), "\n")
cat("Number of columns:", ncol(selected_df), "\n")

# to check number of rows and columns
dim(selected_df)

```

```{r}
# Calculating total number of missing values in the dataset
sum(is.na(selected_df))
```

```{r}
# Counting the number of missing values in each column
colSums(is.na(selected_df))
```

```{r}
# Total percentage of missing values in the dataset
total_missing_percent <- (sum(is.na(selected_df)) / (nrow(selected_df) * ncol(selected_df))) * 100
total_missing_percent
```

```{r}
# Removing rows with any missing values
cleaned_df <- selected_df %>% drop_na()
cleaned_df
```

```{r}
# Checking for duplicate rows
sum(duplicated(cleaned_df))
```

```{r}
# Calculating total number of missing values 
sum(is.na(cleaned_df))
```


```{r}
# To check data types of all columns
str(cleaned_df)
```
```{r}
# Renaming variables name
cleaned_df <- cleaned_df %>%
  rename(
    "Country" = "Country Name",
    "Renewable energy consumption (%)" = "Renewable energy consumption (% of total final energy consumption)",
    "Renewable electricity output (%)" = "Renewable electricity output (% of total electricity output)",
    "Annual production-based emissions of carbon dioxide (in million tonnes)" = "Annual production-based emissions of carbon dioxide (CO2), measured in million tonnes",
    "GDP per capita (US$)" = "GDP per capita (current US$)",
    "Primary school enrollment (% gross)" = "School enrollment, primary (% gross)",
    "c" = "School enrollment, secondary (% gross)",
    "Income Classification" = "Income Classification (World Bank Definition)",
    "Individuals using the Internet (%)" = "Individuals using the Internet (% of population)",
    "Life expectancy (years)" = "Life expectancy at birth, total (years)",
    "Population" = "Population, total",
    "Compulsory education duration (years)" = "Compulsory education, duration (years)"
  )

```

## Exploratory Data Analysis

```{r}
# Provides a statistical summary of Renewable energy consumption
summary(cleaned_df$"Renewable energy consumption (%)")
```

```{r fig.cap="Figure 6.1: Distribution of Renewable Energy Consumption (%)"}
# Plotting the distribution of Renewable Energy Consumption
ggplot(cleaned_df, aes(x = `Renewable energy consumption (%)`)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Renewable Energy Consumption (%)",
    x = "Renewable Energy Consumption (%)",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
# Calculate IQR and bounds
Q1 <- quantile(cleaned_df$`Renewable energy consumption (%)`, 0.25, na.rm = TRUE)
Q3 <- quantile(cleaned_df$`Renewable energy consumption (%)`, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify potential outliers
outliers <- cleaned_df %>%
  filter(`Renewable energy consumption (%)` < lower_bound |
         `Renewable energy consumption (%)` > upper_bound)

# View number of outliers
nrow(outliers)
```

```{r}
# Filter outlier rows
outliers <- cleaned_df %>%
  filter(`Renewable energy consumption (%)` < lower_bound | `Renewable energy consumption (%)` > upper_bound)

# View the countries considered outliers
outliers %>%
  select(Country, `Renewable energy consumption (%)`)

```

```{r}
# Provides a statistical summary of Annual production of  carbon dioxide
summary(cleaned_df$"Annual production-based emissions of carbon dioxide (in million tonnes)")
```

```{r fig.cap="Figure 6.2: Distribution of Annual production-based emissions of carbon dioxide"}
# Plotting the distribution of CO2 emissions across all countries
ggplot(cleaned_df, aes(x = `Annual production-based emissions of carbon dioxide (in million tonnes)`)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of CO2 Emissions (Million Tonnes)",
    x = "CO2 Emissions",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
# Calculating Outliers
Q1 <- quantile(cleaned_df$`Annual production-based emissions of carbon dioxide (in million tonnes)`, 0.25, na.rm = TRUE)
Q3 <- quantile(cleaned_df$`Annual production-based emissions of carbon dioxide (in million tonnes)`, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

co2_outliers <- cleaned_df %>%
  filter(`Annual production-based emissions of carbon dioxide (in million tonnes)` < lower_bound |
         `Annual production-based emissions of carbon dioxide (in million tonnes)` > upper_bound)

# View distinct outlier countries
co2_outliers %>%
  select(Country, Year, `Annual production-based emissions of carbon dioxide (in million tonnes)`)
```

```{r fig.cap="Figure 6.3: Distribution of CO2 Emissions"}
# Removing outliers
cleaned_df %>%
  filter(`Annual production-based emissions of carbon dioxide (in million tonnes)` <= 1000) %>%
  ggplot(aes(x = `Annual production-based emissions of carbon dioxide (in million tonnes)`)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of CO2 Emissions (Million Tonnes) After Removing Outliers",
    x = "CO2 Emissions (Million Tonnes)",
    y = "Frequency"
  ) +
  theme_minimal()

```

```{r}
# Provides a statistical summary of GDP per capita 
summary(cleaned_df$`GDP per capita (US$)`)
```

```{r fig.cap="Figure 6.4: Distribution of GDP per Capita across all countries"}
# Plotting the distribution of GDP per Capita
ggplot(cleaned_df, aes(x = `GDP per capita (US$)`)) +
  geom_histogram(binwidth = 2000, fill = "darkgreen", color = "white") +
  labs(
    title = "Distribution of GDP per Capita (US$)",
    x = "GDP per Capita (US$)",
    y = "Frequency"
  ) +
  theme_minimal()

```

```{r}
# Calculating outliers
Q1 <- quantile(cleaned_df$`GDP per capita (US$)`, 0.25, na.rm = TRUE)
Q3 <- quantile(cleaned_df$`GDP per capita (US$)`, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

gdp_outliers <- cleaned_df %>%
  filter(`GDP per capita (US$)` < lower_bound | `GDP per capita (US$)` > upper_bound)

# View outlier countries
gdp_outliers %>% select(Country, Year, `GDP per capita (US$)`)
```
A total of 116 entries were identified as outliers in GDP per capita. These high outlier values correspond to countries with exceptionally strong economies, such as Luxembourg, United States, etc. Rather than indicating data errors, these figures represent real-world economic extremes and are essential to understanding global income diffrences. As such, the outliers are retained in the dataset to preserve the integrity of the analysis and to accurately reflect the economic diversity among countries.

**D. Secondary school enrollment (% gross)**

* **Step 1:** Summary Statistics

```{r}
# Provides a statistical summary of Secondary school enrollment
summary(cleaned_df$`Secondary school enrollment (% gross)`)
```
The gross secondary school enrollment rate ranges from a minimum of 10.64% to a maximum of 163.93%. The median value is 90.53%, while the mean is slightly lower at 83.52%. The first quartile is 69.60%, and the third quartile is 101.05%, indicating that most countries fall within a fairly wide range of enrollment levels. Also, values exceeding 100% are possible in gross enrollment data, as they may include over-aged or under-aged students who are enrolled at the secondary level.

* **Step 2:** Visualising the Distribution

```{r fig.cap="Figure 6.5: Distribution of Secondary School Enrollment (% Gross)  across all countries"}
ggplot(cleaned_df, aes(x = `Secondary school enrollment (% gross)`)) +
  geom_histogram(binwidth = 5, fill = "purple", color = "white") +
  labs(
    title = "Distribution of Secondary School Enrollment (% Gross)",
    x = "Secondary School Enrollment (% Gross)",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r}
# calculating Outliers
Q1 <- quantile(cleaned_df$`Secondary school enrollment (% gross)`, 0.25, na.rm = TRUE)
Q3 <- quantile(cleaned_df$`Secondary school enrollment (% gross)`, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

enrollment_outliers <- cleaned_df %>%
  filter(`Secondary school enrollment (% gross)` < lower_bound | `Secondary school enrollment (% gross)` > upper_bound)

# View outlier countries
enrollment_outliers %>% select(Country, Year, `Secondary school enrollment (% gross)`)
```

```{r}
# Provides a statistical summary of % of population with access to electricity
summary(cleaned_df$`Access to electricity (% of population)`)
```

```{r fig.cap="Figure 6.6: Distribution of % of population with Access to electricity across all countries"}
ggplot(cleaned_df, aes(x = `Access to electricity (% of population)`)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Access to Electricity (%)",
    x = "Access to Electricity (%)",
    y = "Frequency"
  ) +
  theme_minimal()
```

```{r fig.cap="Figure 7.1: Relationship between Renewable Energy Consumption and CO₂ Emissions across Countries (2010–2015."}
# Plotting the relationship between the two variables
cleaned_df %>%
  filter(
    Year >= 2010 & Year <= 2015,
    `Annual production-based emissions of carbon dioxide (in million tonnes)` <= 1000
  ) %>%
  mutate(Population_millions = Population / 1e6) %>%
  ggplot(aes(
    x = `Renewable energy consumption (%)`,
    y = `Annual production-based emissions of carbon dioxide (in million tonnes)`,
    size = Population_millions,
    fill = Continent
  )) +
  geom_point(shape = 21, alpha = 0.6, color = "black") +
  scale_size(range = c(2, 9), name = "Population (Millions)") +
  scale_fill_manual(
    name = "Continent",
    values = c(
      "Africa" = "green",
      "Asia" = "pink",
      "Europe" = "lightblue",
      "North America" = "grey",
      "Oceania" = "yellow",
      "South America" = "orange"
    )
  ) +
  labs(
    title = "Renewable Energy vs CO2 Emissions (Global Overview, 2010–2015)",
    x = "Renewable Energy Consumption (%)",
    y = "CO2 Emissions (Million Tonnes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )
```

```{r fig.cap="Figure 7.2: Yearly Comparison of Renewable Energy Consumption and CO₂ Emissions by Country (2010–2015)."}
# Filter the dataset with emissions threshold included
bubble_data <- cleaned_df %>%
  filter(
    Year %in% 2010:2015,
    !is.na(`Renewable energy consumption (%)`),
    !is.na(`Annual production-based emissions of carbon dioxide (in million tonnes)`),
    !is.na(Population),
    !is.na(Continent),
    `Annual production-based emissions of carbon dioxide (in million tonnes)` <= 1000
  ) %>%
  mutate(Population_millions = Population / 1e6)

# Plot
ggplot(bubble_data, aes(
  x = `Renewable energy consumption (%)`,
  y = `Annual production-based emissions of carbon dioxide (in million tonnes)`,
  size = Population_millions,
  fill = Continent
)) +
  geom_point(shape = 21, alpha = 0.7, color = "black") +
  facet_wrap(~ Year, ncol = 3) +
  scale_size(range = c(2, 9), name = "Population (Millions)") +
  scale_fill_manual(
    name = "Continent",
    values = c(
      "Africa" = "green",
      "Asia" = "pink",
      "Europe" = "lightblue",
      "North America" = "grey",
      "Oceania" = "yellow",
      "South America" = "orange"
    )
  ) +
  labs(
    title = "Renewable Energy vs CO2 Emissions by Countries (2010–2015)",
    x = "Renewable Energy Consumption (%)",
    y = "CO2 Emissions (Million Tonnes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold")
  )

```

```{r fig.cap="Figure 7.3: Relationship Between Secondary School Enrollment and GDP per Capita by Development Level"}
# Combine mutate + scatter plot
cleaned_df %>%
  mutate(
    Development_Level = case_when(
      `Income Classification` %in% c("High income", "Upper-middle income") ~ "Developed",
      `Income Classification` %in% c("Lower-middle income", "Low income") ~ "Developing",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(
    x = `Secondary school enrollment (% gross)`,
    y = `GDP per capita (US$)`,
    color = Development_Level
  )) +
  geom_point(size = 2, alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(
    values = c("Developed" = "blue", "Developing" = "red"),
    name = "Development Level"
  ) +
  labs(
    title = "Investment in Education vs GDP per Capita",
    subtitle = "Secondary School Enrollment and Economic Growth by Development Level",
    x = "Secondary School Enrollment (% Gross)",
    y = "GDP per Capita (US$)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )


```

```{r fig.cap="Figure 7.4: Top 10 Countries with the Lowest Average Access to Electricity (2008–2018)}
# Filter for years 2009–2019 and compute average access
electricity_avg <- cleaned_df %>%
  filter(Year >= 2008 & Year <= 2018) %>%
  group_by(`Country`) %>%
  summarise(avg_access = mean(`Access to electricity (% of population)`, na.rm = TRUE)) %>%
  arrange(avg_access) %>%
  slice(1:10)

# Convert country to factor to preserve order in plot
electricity_avg <- electricity_avg %>%
  mutate(`Country` = factor(`Country`, levels = rev(`Country`)))

# Plot
ggplot(electricity_avg, aes(x = avg_access, y = `Country`)) +
  geom_bar(stat = "identity", fill = "blue", width = 0.6) +
  geom_text(aes(label = paste0(round(avg_access, 1), "%")),
            hjust = -0.1,
            size = 4.2) +
  scale_x_continuous(limits = c(0, 35), expand = c(0, 0)) +
  labs(
    title = "Countries with the Lowest Access to Electricity",
    subtitle = "Based on Average Access Over 2008-2018 Period",
    x = "Avg. Access to Electricity (%)",
    y = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = 11),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11, margin = margin(b = 10))
  )
```

```{r fig.cap="Figure 7.5: Electricity Access vs GDP per Capita in 2015}
# Filter the data for the selected year (e.g., 2018)
electricity_data <- cleaned_df %>%
  filter(Year == 2015, 
         !is.na(`Access to electricity (% of population)`),
         !is.na(`GDP per capita (US$)`))

# Create scatter plot with editable colors
ggplot(electricity_data, aes(x = `Access to electricity (% of population)`, 
                             y = `GDP per capita (US$)`, 
                             color = Development_Level)) +
  geom_jitter(width = 1, height = 0, alpha = 0.8, size = 2) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(
    values = c(
      "Developed" = "blue",     # Change to any hex or color name
      "Developing" = "red"     # Change to any hex or color name
    )
  ) +
  labs(
    title = "Electricity Access vs GDP per Capita",
    subtitle = "2015: Developed vs Developing Countries",
    x = "Access to Electricity (%)",
    y = "GDP per Capita (US$)",
    color = "Development Level"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

```

## Conclusion
This project focused on analysing global sustainability indicators across countries between 2000 and 2018 using R. The analysis involved cleaning and transforming the dataset, handling missing values, and selecting key economic, environmental, and social variables. Exploratory analysis and visualisations were then used to uncover global patterns and relationships among indicators such as renewable energy use, CO₂ emissions, GDP growth, and access to electricity.  

The results showed clear disparities in sustainability progress across regions. Developed economies generally exhibited higher GDP growth and electricity access but also higher CO₂ emissions, while developing nations showed gradual improvements in renewable energy use. Overall, this project highlights the value of using data-driven approaches to evaluate global sustainability efforts and provides a foundation for further exploration into factors influencing environmental and economic development worldwide.

